Urbanization affects web abundance and aggregation of a funnel-weaving spider, *Agelenopsis pennsylvanica*

### 2023-01-08 manuscript/manuscript.Rmd
Lots of changes

---
title: "Urbanization affects web abundance and aggregation of a funnel weaver, *Agelenopsis pennsylvanica*"
author: Brandi Pessman, Madison Hays, Earl Agpawa, Eileen Hebets
date: "`r Sys.Date()`"
output: 
  bookdown::word_document2:
    reference_docx: word_styles_reference_01.docx
    fig_caption: true
    number_sections: false
bibliography: [citations.bib]
csl: apa.csl
---

@@

**Log line: Under increasing urbanization, generalist predators actively select web locations to effectively capitalize on more abundant prey as environmental conditions change.**

@@

Animals distribute themselves within habitats based on a variety of environmental conditions, including those impacted by urbanization. In Lincoln, Nebraska, USA, funnel-weaving spiders (*Agelenopsis pennsylvanica*) are prevalent across urban habitats and actively choose web sites. We compared *A. pennsylvanica* abundance and distribution between two distinct urban habitats: an urban center and an urban forest. We searched along paths from randomly selected start sites in each habitat until we found the first occupied (focal) web. We measured web and spider abundance, distances between the focal and two nearest-neighbor webs, and web height in a ten-meter radius. We found that (i) search distances to focal webs were shorter, and both (ii) webs and (iii) spiders were more abundant in the urban center than in the urban forest, despite less diverse plants for web structures in the urban center. We also found (iv) shorter distances between webs and (v) lower web heights in the urban center. Our data suggest that *A. pennsylvanica* spatially distribute themselves differently across different urban habitats, and we discuss a variety of environmental variables that may predict this behavior.

@@

Animal abundance and distribution are influenced by combinations of environmental factors, intra- and interspecific interactions, and dispersal capacity [@svamberkova_leps_2020; @diez_pulliam_2007]. Environmental factors -- including climatic conditions, land cover, habitat heterogeneity, and access to certain resources (like water and shelter) -- are suspected to relate to physiological constraints that limit the fundamental niche [@chase_leibold_2003]. Interactions within species (between conspecific competitors and mates) and across species (with predators, prey, and heterospecific competitors) further shape animal spatial patterns [@anderson_2017; @boulangeat_2012]. While dispersal capacity can determine a species’ range, barriers can limit dispersal and prevent animals from reaching suitable habitats  [@boulangeat_2012; @amundrud_2018].

Aggregations can form when suitable conditions and resources are not evenly distributed, leading to higher-quality patches [@churchill_2020]. For example, gaps in the tree canopy allow more sunlight to reach the understory -- thus, increasing plant height and soil moisture and decreasing leaf litter -- causing some invertebrate families to become more abundant in these patches [@perry_2018]. Clumping can also result from incomplete dispersal that occurs when barriers, often due to habitat structure, prevent adequate dispersal [@riechert_1974b]. The relative importance of different spatial predictors can vary by the type of taxon-specific spatial information being collected -- *i.e.* presence, abundance, aggregation, or vertical position [@boulangeat_2012].

Due to intense habitat disturbance and dense human presence, urbanization has drastically modified the land [@alberti_2020; @mcdonnell_pickett_1990; @pickett_2001] which has altered environmental conditions, community dynamics, and dispersal capability as a result [@piano_2020]. As these factors determine animal spatial patterns, it is likely that urbanized landscapes have changed how animals spatially distribute themselves, in turn. Typically, city tree cover is reduced in areas where the potential natural vegetation is forested [@nowak_1996] and greater in areas with prairie, grassland, and desert origins [@mckinney_2006]. Urban environments are additionally characterized by decreased biodiversity [@mckinney_2002] and increased impervious cover (compacted soil from buildings and pavement; @nowak_greenfield_2020) and habitat patchiness [@cadenasso_2007]. These conditions can contribute to changes in climatic conditions, such as when impervious cover traps heat and increases the average temperatures in cities compared to surrounding areas –- *i.e.*, the Urban Heat Island Effect [@morabito_2021].

Disturbance from human presence can also lead to air, water, and light pollution [@grimm_2008; @hopkins_2018], as well as air-borne, water-borne, and substrate-borne noise [@barber_2010; @slabbekoorn_2010; @wu_elias_2014]. There is growing evidence that urbanization changes species interactions, even reshaping whole food webs [@start_2020] and the landscape of parasite and disease transmission [@bradley_altizer_2007]. Further, habitat fragmentation and urban features often limit dispersal, which can cause higher degrees of aggregation for urban animals [@johnson_munshisouth_2017]. Urban ecosystems are extremely heterogeneous with unique and diverse habitats nested within [@rivkin_2019]. Habitats within urban ecosystems can vary considerably in feature, condition, and disturbance, resulting in variable access and distributions of resources that can differentially drive animal spatial patterns [@murray_2019; @trubl_2012].

Variation in the abundance and distribution of urban arthropods can have implications for ecosystem function due to the essential ecosystems services that arthropods provide, like pest control [@otoshi_2015] and pollination [@ivers_2022]. Arthropods are considered critical bioindicators of ecosystem health as their short generation time requires them to respond rapidly to human-induced change [@mcintyre_2000]. However, the impact of urbanization on arthropod ecology is still understudied [but see @kotze_2011; @fenoglio_2020], and the relationship between urban habitats and arthropod abundance and diversity remain controversial [@gippet_2017; @dale_frank_2018]. Such discrepancies are probably due to abundance and distributions being species- [@gippet_2017], habitat- [@bennett_lovell_2014], and scale-dependent [@philpott_2014]. While some animals benefit from urban conditions [@guenard_2015], many cannot survive the disturbance and change associated with cities [@wagner_2020]. Thus, it is important to determine how urban factors may be contributing to variation in arthropod abundance and spatial patterns for the conservation of arthropods and their essential services.

Specifically, spiders play an important role in pest control and are predicted to collectively consume up to 800 million tons of arthropods each year [@nyffeler_birkhofer_2017]. Many web-building spiders are choosy in their web locations based on the distribution and abundance of other arthropod prey [@trubl_2012; @riechert_tracy_1975]. Territoriality is a form of intra-specific interactions that can also shape spatial distributions of web-builders [@riechert_1974b]. Further, previous studies have found environmental factors that vary across urban habitats may drive differences in spider spatial distribution, including noise [@bunkley_2017; @gomes_2020], pollution [@zmudzki_laskowski_2012], fragmentation [@stefani_delclaro_2015], microclimate conditions [@riechert_tracy_1975], vegetation structure [@peng_2020], and suitable web substrates [@chan_2009].

The mechanisms behind solitary spider aggregations - especially in urban habitats - are unclear despite reports of aggregations of spiders in urban habitats, such as web-builders *Lactrodectus hesperus* and *Brigittea civica* [@johnson_et_al_2012; @mammola_et_al_2018]. This gap in knowledge is concerning as urban spider aggregations can have medical and aesthetic consequences, and implications for urban pest control services. Prey abundance seems to be an important driver because it has been shown to be directly and indirectly (for example, light sources at night as cues of prey clustering) related to urban spider aggregations (Table 1). Random spacing in solitary spiders is typically maintained by territoriality and aggression over resources like prey and mates [@riechert_1974b]. Occasionally, conditions arise that allow solitary spiders to become more tolerant of neighbors, and understanding these contexts can help uncover the mechanisms of aggregative behavior. Aggregations have been shown to form as a result of inter- and intraspecific interactions, environmental factors, and dispersal limitations (Table 1).

Funnel weavers in the genus *Agelenopsis* actively select web locations based on experience of perceived quality of the site [@riechert_1976], and are thus likely impacted by environmental conditions resulting from urbanization. Prey abundance, available web substrates, vegetation structure, and microclimate conditions have been shown to be indicators of a quality web site in a desert habitat for *A. aperta* [@riechert_1974; @riechert_1973]. *Agelenopsis pennsylvanica* build webs that consist of a sheet that functions as a prey capture area and a retreat to escape from predators and unfavorable conditions (Fig. 1), like direct sunlight [@riechert_tracy_1975]. The horizontal sheet is not sticky, and the web is used solely for its vibratory properties to rapidly and accurately locate prey that fall upon or walk across the web [@singer_2000]. Occasionally, vertical silk strands are found that may function to capture flying prey or evade predators [@singer_2000]. The spider sits in the retreat until the prey is sensed and returns to the retreat immediately after capturing the prey. Juvenile males and females build semi-permanent webs, but males leave their web shortly after maturity in search of a receptive female [@singer_2000].

Understanding how urban spiders are affected by changes in environmental conditions and habitat structure brought on by processes of urbanization is vital as cities continue to densify and expand. Thus, we compared *A. pennsylvanica* spatial patterns between two urban habitats (an urban forest and an urban center) and explored what environmental factors are important determinants of web placement. In particular, we use *A. pennsylvanica* to address three questions: (*i*) How do web abundance, aggregation, and position vary between two urban habitats? (*ii*) How do environmental conditions vary between these two urban habitats? (*iii*) What environmental variables help predict spider spatial patterns in distinct urban habitats? Based on preliminary observations, we predicted that spiders would be more abundant and aggregated in the urban center than the urban forest. We further hypothesized that environmental conditions that vary between the two habitats would be correlated with the variation in the spatial patterns of *A. pennsylvanica*.

@@

### Urban Habitats

We conducted our study in September and October of 2020 in Lincoln, Nebraska, USA. Lincoln -- with a population of about 300,000 people [@census] -- has diverse urban habitats that vary in the degree of disturbance, management, and plant cover. We compared the abundance, aggregation, and position of funnel webs (*Agelenopsis pennsylvanica*) between two urban habitats: the University of Nebraska-Lincoln (UNL) City Campus (an urban center) and Wilderness Park (an urban forest) (Fig. 2, Fig. S1). These two habitats were selected due to their distinctive characteristics (described below), preliminary observations of potential differences in funnel weaver abundance and aggregation (Fig. S2), and ease of access.

To first quantify differences in the environmental conditions between urban habitats, we gathered and analyzed Geographic Information Systems (GIS) data in QGIS (v. 3.16.3-Hannover, ESRI 102704). We obtained tree cover [@usfs_2016], impervious (building and pavement) cover [@usgs_2019_imperv], and land use class [@usgs_2019_land] from the National Land Cover Database (30-meter resolution). We used an image from path 28, row 32 of the Landsat-7 World Reference System-2 (Band 6) to collect spectral radiance -- a measure of thermal conditions on Earth's surface (30-meter resolution, taken on September 2, 2020). Finally, we assessed artificial light at night levels using the 2020 Visible Infrared Imaging Radiometer Suite (VIIRS) from https://www.lightpollutionmap.info/ (500-meter resolution). We polygonized the tiff files and calculated the average value for each factor by location: UNL City Campus, Wilderness Park, and Lincoln City Limits (see Fig. 2, Fig. S3-S6). We included measurements for Lincoln city limits to show how the two study locations compare to the city of Lincoln overall. We calculated the averages using the following equation:

@@

Where *i* is the value indicated by the pixel (with range b-a) and *$x_i$* is the percent of the habitat covered by the *$i^(th)$* pixel category. These data are presented in Table 2 and summarized below.

We defined the study boundaries of UNL City Campus as Salt Creek Roadway, North Antelope Valley Parkway, North 10th Street, and Q Street (1.12 $km^2$, Fig. 2, Fig. S1). UNL City Campus is adjacent to, and just north of, downtown Lincoln (253.73 $km^2$). Impervious cover -- in the form of buildings and pavement -- dominates the land on UNL City Campus, considerably more than impervious cover across the city of Lincoln (Table 2, Fig. S3). Lawns are maintained by frequent mowing, and gardens and landscaping are regularly pruned. Tree cover on UNL City Campus and in Lincoln overall is low (< 6%, Table 2, Fig. S4). There was heavy foot traffic from UNL's 20,286 undergraduates enrolled in the 2020-2021 academic year, 37% of which lived in college-owned, -operated, or -affiliated housing [@iea_2021]. The entire area of UNL City Campus was categorized as urban with varying intensity (Table 2, Fig. 2). Over half of UNL City Campus was high intensity urban cover compared to only 13.16% for Lincoln overall. Further, UNL City Campus exhibits heighten spectral radiance (thermal conditions, Fig. S5) and artificial light at night radiance (Fig. S6), even above that of Lincoln as a whole (Table 2). Thus, we designated UNL City Campus as an *urban center* and refer to it as such hereafter.

Wilderness Park is a stretch of forested land (6.44 $km^2$, Fig. 2, Fig. S1) starting about 3.58 km southwest of UNL City Campus and downtown Lincoln, extending south and along Lincoln city limits. We found about two-fifths of Wilderness Park to be shrouded in trees (Table 2, Fig. S4), with less than 2% of its land imperviously covered (Table 2, Fig. S2). Over three quarters of Wilderness Park is categorized as forested land with only 5.44% labeled as urban (Table 2, Fig. 2). Wilderness Park has about 18 kilometers of dirt trails, dense deciduous tree growth, and variable undergrowth. The trails are open for recreational walking, running, biking, and horse riding. While we found no public records of the number of visitors to Wilderness Park, trail-goers were infrequent during data collection. Wilderness Park showed greatly reduced spectral radiance (Fig. S5) and artificial light at night radiance (Fig. S6) than the city of Lincoln (Table 2). Due to its proximity to Lincoln city limits, we designate Wilderness Park as an *urban forest* and refer to it as such hereafter.

### Study Sites and Field Measurements

Within the boundaries of the urban forest, we assigned numbers to intersections of three or more trails (Fig. S7). Similarly, we assigned numbers to each intersection of roads and sidewalks that meet from the four cardinal directions in the urban center (Fig. S7).  We used a random number generator to randomly select 10 intersections in the urban forest and 12 intersections in the urban center as start sites (Table S1). We marked the GPS coordinates of these locations using a GPS receiver (Magellan SporTrak Map, Model 980616-01, Parsippany, New Jersey, USA).

From each start site, we began walking along a path in a randomly selected direction (North, East, South, or West using a random number generator). We thoroughly searched the edges of the path for funnel webs as the GPS receiver tracked our movement (Fig. S8). We walked along pre-existing sidewalks, roads, and trails because paths were a common feature of both habitats along which we had previously found funnel weavers. We often find *A. pennsylvanica* in edge habitats in the study region, and this tendency to build webs at a habitat's edge is shared by other spider species [@cobbold_supp_2012]. Additionally, the thick undergrowth of the urban forest often restricted searches off the trails. To ensure equivalent search efforts, we took more time searching the edge of dense undergrowth in the urban forest.

When we reached a funnel web, we checked to see if the web was occupied by a spider by resting a toothpick secured to the bristled end of a vibrating toothbrush on the sheet of the web to simulate prey in the web. If a spider emerged, we noted this web as the *focal web*. If no spider emerged after 30 seconds, we continued searching until we found an occupied funnel web. We used the GPS receiver to mark the coordinates of the focal web. Although it is possible that other species of *Agelenopsis* are present in Lincoln, we have only found *A. pennsylvanica* during surveys in the area.

Search distance can often be used to estimate abundance [@buckland_2005], as spiders are likely less abundant in an area where we need to search further from the start site to the focal web. Thus, we use search distance as evidence for variation in abundance. To quantitatively determine the abundance of webs and spiders near each focal web, we measured a ten-meter radius around the focal web and counted the number of occupied and unoccupied webs determined in the same manner as the focal web. We also assessed nearest neighbor distances as a measure of aggregation behavior. Starting at the focal web and working outwards, we marked the locations of the two webs (including occupied and unoccupied) nearest to the focal web within ten meters. We limited our measurements to the nearest two webs because we often found no more than three webs (including the focal web) per site in the urban forest. We measured the shortest distance from the retreat of the focal web to the retreat of the two nearest neighbor webs. To explore differences in web placement, we measured the web height from the ground to the web retreat for every web with the ten-meter radius. Differences in web height have been shown to vary by environmental factors associated with urbanization [@dahirel_2017]. To get an idea of plant species richness at each site, we recorded each unique plant species found in a ten-meter radius of the focal web (Fig. S8).

We collected data from September 23 to October 14, 2020. During these dates, all *A. pennsylvanica* have reached maturity [@guarisco_2014; personal observation]. Of the 64 occupied webs, 61 were occupied by females indicating that males had likely abandoned their webs in search of females. The 67 unoccupied webs could be those once owned by males or those abandoned by females. While webs were found on many of the same plant types between sites in the urban forest and urban center in varying proportions, 15% of webs in urban center sites were built on non-plant substrates (Fig. S9).

### Environmental Predictors

To assess potential mechanisms behind variation in abundance, aggregation, and position, we looked for correlations between spider spatial patterns and environmental conditions -- collected in the field (plant species richness) or using GIS for each site. In QGIS, we created 100 meter buffers around each focal web location and processed the GIS variables mentioned above (tree cover, impervious cover, spectral radiance, and artificial light at night radiance) in the same manner as before. To these we added patch area, potential disturbance from nearby roads and highways, and road length in the 100 meter buffer. We used a 100 meter buffer to accommodate the resolution of the GIS data. We refrained from including a larger buffer region as the dispersal range of this system is unknown and may be dependent on barriers present in the habitat.

>>>>

---
title: "Urbanization affects web abundance and aggregation of a funnel-weaving spider, *Agelenopsis pennsylvanica*"
output: 
  bookdown::word_document2:
    reference_docx: word_styles_reference_01.docx
    fig_caption: true
    number_sections: false
bibliography: [citations.bib]
csl: apa.csl
---

@@

Brandi J. Pessman ^1\*^, Madison Hays ^1^, Earl Agpawa ^2^, Eileen A. Hebets ^1^

^1^ University of Nebraska-Lincoln, Department of Biological Sciences, Lincoln, Nebraska, United States

^2^ University of Nebraska-Lincoln, Department of Entomology, Lincoln, Nebraska, United States

^\*^ Corresponding Author Email: [bjpessman\@gmail.com](mailto:bjpessman@gmail.com){.email}

@@

Animals distribute themselves within habitats based on a variety of environmental conditions, including those impacted by urbanization. Throughout North America, funnel-weaving spiders (*Agelenopsis*) are prevalent across urban habitats and actively choose sites to build webs. We compared *A. pennsylvanica* abundance and distribution between two distinct urban habitats: an urban center (university campus) and an urban forest (city park). These urban habitats differ significantly in terms of plant diversity and disturbance from nearby roads and highways, among other factors. We wanted to know how these urban habitats and their associated environmental conditions affect *A. pennsylvanica* spatial patterns. We searched along paths from randomly selected start sites in each habitat until we found the first occupied (focal) web. We measured web and spider abundance, distances between the focal and two nearest-neighbor webs, and web height in a ten-meter radius. We found that (i) search distances to focal webs were shorter, and both (ii) webs and (iii) spiders were more abundant in the urban center than in the urban forest, despite less diverse plants for web structures in the urban center. We also found (iv) shorter distances between webs and (v) lower web heights in the urban center. Generally, spiders were more abundant and aggregated at sites that had lower plant diversity and greater disturbance from road and highway traffic. Our data suggest that *A. pennsylvanica* spatially distribute themselves differently across different urban habitats, and this variation relates, in part, to differences in environmental conditions.

## Keywords
Arachnids, Arthropods, Spatial Distribution, Environmental Predictors

## Acknowledgments

We thank Morgan Ryan for his advice about processing the GIS data in QGIS. We are grateful for past and present members of the Hebets, Shizuka, and Keesey labs (UNL Behavior Group) for their feedback on data analysis and earlier drafts of the manuscript. We also thank conference attendees that have visited posters and oral talks about this project, as those conversations have helped to improve many aspects of the research presented here.

\newpage

@@

Animal abundance and distribution are influenced by combinations of environmental factors, intra- and interspecific interactions, and dispersal capacity [@svamberkova_leps_2020; @diez_pulliam_2007]. Environmental factors, including climatic conditions, land cover, habitat heterogeneity, and access to certain resources (like water and shelter), are suspected to relate to physiological constraints that limit the fundamental niche [@chase_leibold_2003]. Interactions within species (between conspecific competitors and mates) and across species (with predators, prey, and heterospecific competitors) further shape animal spatial patterns [@anderson_2017; @boulangeat_2012]. While dispersal capacity can determine a species' range, barriers can limit dispersal and prevent animals from reaching suitable habitats [@boulangeat_2012; @amundrud_2018].

Aggregations can form when suitable conditions and resources are not evenly distributed, leading to higher-quality patches [@churchill_2020]. For example, gaps in the tree canopy allow more sunlight to reach the understory -- thus, increasing plant height and soil moisture and decreasing leaf litter -- causing some invertebrate families to become more abundant in these patches [@perry_2018]. Clumping can also result from incomplete dispersal that occurs when barriers, often due to habitat structure, prevent adequate dispersal [@riechert_1974b]. Scientists have found differences in the relative importance of different spatial predictors based on the type of taxon-specific spatial information collected -- *i.e.* presence, abundance, aggregation, or vertical position [@boulangeat_2012].

Due to intense habitat disturbance and dense human presence, urbanization has drastically modified the land [@alberti_2020; @mcdonnell_pickett_1990; @pickett_2001]. These land changes lead to altered environmental conditions, community dynamics, and dispersal capability as a result [@piano_2020]. Urban environments are characterized by decreased biodiversity [@mckinney_2002], increased impervious cover [compacted soil from buildings and pavement, @nowak_greenfield_2020], habitat patchiness [@cadenasso_2007], pollution [air, water, and light, @grimm_2008; @hopkins_2018], and noise [air-, water-, and substrate-borne, @barber_2010; @slabbekoorn_2010; @wu_elias_2014]. These conditions can contribute to changes in local climatic conditions, such as when impervious cover traps heat and increases the average temperatures in cities compared to surrounding areas --- *i.e.*, the Urban Heat Island Effect [@morabito_2021]. There is growing evidence that urbanization changes species interactions, even reshaping whole food webs [@start_2020] and altering the landscape of parasite and disease transmission [@bradley_altizer_2007]. As many of these conditions and dynamics determine animal spatial patterns, it is likely that urbanized landscapes have changed how animals spatially distribute themselves.

Urban ecosystems can also be extremely heterogeneous with unique and diverse habitats nested within [@rivkin_2019]. For example, habitats within urban ecosystems can differ considerably in landscape features (e.g. tree and building cover), weather conditions (e.g. temperature, humidity, and wind speed), and disturbance, resulting in variable access and distributions of resources and opportunities for certain species to thrive. These factors can also differentially drive animal spatial patterns [@murray_2019; @trubl_2012], sometimes leading to higher degrees of aggregation for urban animals [@johnson_munshisouth_2017].

Variation in the abundance and distribution of urban arthropods can have implications for ecosystem function due to the essential ecosystem services that arthropods provide, like pest control [@otoshi_2015] and pollination [@ivers_2022]. Arthropods are considered critical bioindicators of ecosystem health as their short generation time requires them to respond rapidly to human-induced change [@mcintyre_2000]. However, the impact of urbanization on arthropod ecology is still understudied [but see @kotze_2011; @fenoglio_2020], and the relationship between urban habitats and arthropod abundance and diversity remain controversial [@gippet_2017; @dale_frank_2018]. Such discrepancies are most likely due to abundance and distributions being species- [@gippet_2017], habitat- [@bennett_lovell_2014], and scale-dependent [@philpott_2014]. While some animals benefit from urban conditions [@guenard_2015], many cannot survive the disturbance and change associated with cities [@wagner_2020]. Thus, it is important to determine how urban factors contribute to variation in arthropod abundance and spatial patterns for the conservation of arthropods and their essential services.

Spiders, specifically, play an important role in pest control. They are predicted to collectively consume up to 800 million tons of arthropods each year [@nyffeler_birkhofer_2017]. Previous studies have found that factors that vary across urban habitats may drive differences in spider spatial distribution, including prey abundance [@trubl_2012; @riechert_tracy_1975], noise [@bunkley_2017; @gomes_2020], pollution [@zmudzki_laskowski_2012], fragmentation [@stefani_delclaro_2015], microclimate conditions [@riechert_tracy_1975], vegetation structure [@peng_2020], and suitable web substrates [@chan_2009]. Random spacing in solitary spiders is typically maintained by territoriality and aggression over resources like prey and mates [@riechert_1974b], but occasionally, conditions arise that allow solitary spiders to become more tolerant of neighbors. The mechanisms behind non-social spider aggregations - especially in urban habitats - are unclear. Nonetheless, there are reports of aggregations of spiders in urban habitats, such as the web-building spiders *Lactrodectus hesperus* and *Brigittea civica* [@johnson_et_al_2012; @mammola_et_al_2018]. A more thorough understanding of how and why various species of spiders might distribute themselves differently in urban settings is important, as urban spider aggregations can have medical and aesthetic consequences, as well as implications for urban pest control services.

Funnel-weaving spiders in the genus *Agelenopsis* are found across North America and can be locally very common. They actively select web locations based on experience of perceived quality of the site [@riechert_1976] and are thus likely impacted by environmental conditions resulting from urbanization. Prey abundance, available web substrates, vegetation structure, and microclimate conditions have been shown to be indicators of a quality web site in a desert habitat for *A. aperta* [@riechert_1974; @riechert_1973]. *Agelenopsis pennsylvanica* is a non-desert dwelling relative that similarly builds webs consisting of a sheet that functions as a prey capture area and a retreat to escape from predators and unfavorable conditions (Fig. 1), such as direct sunlight [@riechert_tracy_1975]. The horizontal sheet is not sticky, and the web is used solely for its vibratory properties to rapidly and accurately alert the spider to prey that fall upon or walk across the web [@singer_2000].

```{r spider, dpi = 300}
knitr::include_graphics("/Users/bjpessman/Documents/phd_research_code/Agelenopsis_aggregation/manuscript/figures/spider.png")
```

**Fig. 1** Study System - *Agelenopsis pennsylvanica* (a). The web consists of a retreat where the spider waits for prey to land, fall, or walk on the web sheet (b)

To understand habitat characteristics that might influence the distribution of *A pennsylvanica* across urban environments, we compared *A. pennsylvanica* spatial patterns between two urban habitats (an urban center and an urban forest). In particular, we asked the following three questions: (*i*) How do environmental conditions vary between two urban habitats? (*ii*) How do *A. pennsylvanica* web abundance, aggregation, and position vary between these two urban habitats? (*iii*) What environmental variables help predict spider spatial patterns in distinct urban habitats? Based on preliminary observations, we predicted that *A. pennsylvanica* would be more abundant and aggregated in the urban center than the urban forest. We further expected that specific environmental conditions that vary between the two habitats (e.g. tree cover, plant species richness, and disturbance from roads) would be correlated with the variation in the spatial patterns of *A. pennsylvanica*.

@@

We conducted our study in September and October of 2020 in Lincoln, Nebraska, USA. Lincoln -- with a population of about 300,000 people [@census] -- has diverse urban habitats that vary in the degree of disturbance, management, and plant cover. We compared the abundance, aggregation, and position of funnel webs of the spider *Agelenopsis pennsylvanica* between two urban habitats: the University of Nebraska-Lincoln (UNL) City Campus (an urban center) and Wilderness Park (an urban forest) (Fig. 2, Fig. S1). These two habitats were selected due to their distinctive characteristics (described below), preliminary observations of potential differences in funnel-weaving spider abundance and aggregation (Fig. S2), and ease of access.

*Urban Center* - We defined the study boundaries of UNL City Campus as Salt Creek Roadway, North Antelope Valley Parkway, North 10th Street, and Q Street (1.12 $km^2$, Fig. 2, Fig. S1). UNL City Campus is adjacent to, and just north of, downtown Lincoln (253.73 $km^2$). Lawns are maintained by frequent mowing, and gardens and landscaping are regularly pruned. There was heavy foot traffic from UNL's 20,286 undergraduates enrolled in the 2020-2021 academic year, 37% of which lived in college-owned, -operated, or -affiliated housing [@iea_2021].

*Urban Forest* - Wilderness Park is a stretch of forested land (6.44 $km^2$, Fig. 2, Fig. S1) starting about 3.58 km southwest of UNL City Campus and downtown Lincoln, extending south and along Lincoln city limits. Wilderness Park has about 18 kilometers of dirt trails, dense deciduous tree growth, and variable undergrowth. The trails are open for recreational walking, running, biking, and horse riding. While we found no public records of the number of visitors to Wilderness Park, trail-goers were infrequent during data collection.

```{r land_location, fig.width = 6.85, dpi = 300}
knitr::include_graphics("/Users/bjpessman/Documents/phd_research_code/Agelenopsis_aggregation/manuscript/figures/land_locations.png")
```

**Fig. 2** Land cover raster with borders for Lincoln, Nebraska, USA city limits (dashed border), University of Nebraska-Lincoln (UNL) City Campus (Urban Center, solid border, a), and Wilderness Park (Urban Forest, solid border, b) using the 2019 National Land Cover Data on Land Cover and QGIS (v. 3.16.3-Hannover, ESRI 102704). For visualization and listing of study sites, see Figure S1 and Table S1, respectively

### Study sites

To select random survey sites in each of the two habitats, we assigned numbers to each intersection of roads and sidewalks that meet from the four cardinal directions in the urban center and to intersections of three or more trails in the urban forest (Fig. S3). We used a random number generator to randomly select 12 intersections in the urban center and 10 intersections in the urban forest as start sites (Table S1). We marked the GPS coordinates of these locations using a GPS receiver (Magellan SporTrak Map, Model 980616-01, Parsippany, New Jersey, USA).

From each start site, we began walking along a path in a randomly selected direction (North, East, South, or West using a random number generator). We conducted a thorough visual search of the edges of the path for *A. pennsylvanica* funnel webs as the GPS receiver tracked our movement (Fig. S4). We walked along pre-existing sidewalks, roads, and trails because paths were a common feature of both habitats along which we had previously found funnel-weaving spiders. We often found *A. pennsylvanica* in edge habitats in the study region, and this tendency to build webs at a habitat's edge is shared by other spider species [@cobbold_supp_2012]. Additionally, the thick undergrowth of the urban forest often restricted searches off the trails. To ensure equivalent search efforts, we took more time searching the edge of dense undergrowth in the urban forest.

*Focal web* -- When we reached a funnel web, we checked to see if the web was occupied by a spider. We did this by resting a toothpick secured to the bristled end of a vibrating toothbrush on the sheet of the web. We turned on the electric toothbrush to simulate prey in the web. If a spider emerged, we noted this web as the *focal web*. If no spider emerged after 30 seconds, we continued searching until we found an occupied funnel web. Preliminary trials demonstrated this to be an effective means of enticing a spider out of its retreat and onto its sheet web. We used the GPS receiver to mark the coordinates of the focal web. Although it is possible that other species of *Agelenopsis* are present in Lincoln, we have only found *A. pennsylvanica* during surveys in the area.

### Urban habitat characteristics

*Environmental conditions of the habitats and Lincoln* -- We first compared differences in environmental conditions across our two selected urban habitats and Lincoln city limits. We chose tree cover, impervious cover, and land cover class for this initial analysis to capture the degree of variation between the habitats as compared to the entire city. We gathered and analyzed Geographic Information Systems (GIS) data in QGIS (v. 3.16.3-Hannover, ESRI 102704). We obtained tree cover [@usfs_2016], impervious cover [building and pavement, @usgs_2019_imperv], and land cover class [@usgs_2019_land] from the National Land Cover Database (30-meter resolution). We polygonized the tiff files and calculated the average value for each factor by location using the boundaries described above: urban center (UNL City Campus), urban forest (Wilderness Park), and Lincoln City Limits (see Fig. 2, Fig. S5-S8). We calculated the averages using the following equation:

@@

Where *i* is the value indicated by the pixel (with range b-a) and $x_i$ is the percent of the habitat covered by the $i^(th)$ pixel category.

*Environmental variables between habitats* -- To add environmental variables for a more detailed comparison of these two habitats, we collected nine total environmental predictors from each study site: (1) tree cover, (2) impervious cover (percent developed surface), (3) thermal conditions (spectral radiance), (4) light at night radiance, (5) plant species richness, (6) patch area (area of continuous vegetation), (7) disturbance from roads, (8) disturbance from highways, and (9) total road length. All measurements except plant species richness were completed in QGIS (v. 3.16.3-Hannover, ESRI 102704).

We recorded the percent tree cover (1) and impervious cover (2) at each site using the polyganized tiff files described above. To collect spectral radiance (3) -- a measure of thermal conditions on Earth's surface, we used an image from path 28, row 32 of the Landsat-7 World Reference System-2 (Band 6, 30-meter resolution, taken on September 2, 2020). We assessed artificial light at night levels (4) using the 2020 Visible Infrared Imaging Radiometer Suite (VIIRS) from <https://www.lightpollutionmap.info/> (500-meter resolution). We analyzed these measures in 100-meter buffers around each focal web. We used a 100-meter buffer to accommodate the resolution of the GIS data. We refrained from including a larger buffer region as the dispersal range of this system is unknown and may be dependent on barriers present in the habitat.

We collected plant species richness (5) on-site by recording each unique plant species (using a plant identification phone application - PictureThis) found in a ten-meter radius of the focal web at each study site (Fig. S9). We measured patch area (6) using satellite imagery to define the area of continuous vegetation at each study site. Roads, sidewalks, and trails were defined as fragment boundaries. To determine the potential disturbance from road (7) and highway traffic (8), we first measured the shortest distance from each focal web to the nearest road [GIS vector from: @centerlines] and highway [GIS vector from: @highways] using the NNJoin Plugin [@nnjoin] for QGIS. We used the 2021 Average Daily Traffic Volume report [@adtv_2021] to identify the average daily number of vehicles to travel on each nearest road and highway. We divided the average daily traffic by the shortest distance from the focal web to the nearest road and highway [as in @dooley_2020] for metrics of road disturbance and highway disturbance, respectively. Lastly, we measured the total length of road [9, @centerlines] within a 100-meter buffer of each study site.

### Spider web abundance, patterns, and positions

To determine differences in spider web abundance, aggregation, and position between the two urban habitats, we completed each of the following measurements at each study site within a ten-meter radius of the focal web.

*Spider Abundance* -- Search distance can often be used to estimate abundance [@buckland_2005], as spiders are likely less abundant in an area where we need to search further from the path intersection to the focal web. Thus, we use search distance as evidence for variation in abundance. To quantitatively determine the abundance of webs and spiders near each focal web, we measured a ten-meter radius around the focal web and counted the number of occupied and unoccupied webs. We used the same methods (using the electric toothbrush) as those used for determining the presence/absence of the focal spider.

*Spider Aggregation* -- We assessed nearest neighbor distances as a measure of aggregation behavior. Starting at the focal web and working outwards, we marked the locations of the two webs (including occupied and unoccupied) nearest to the focal web and within ten meters. We limited our measurements to the nearest two webs because we often found no more than three webs (including the focal web) per site in the urban forest. We measured the shortest distance from the retreat of the focal web to the retreat of the two nearest neighbor webs.

To determine whether nearness of neighbors in the urban center was the result of aggregation or increased density from higher abundance, we completed calculations described by Clark and Evans [@clark_evans_1954]. This calculation compares the measured nearest neighbor distance to the expected distance if animals were evenly spaced within an area. For each site, we estimated the total search area by adding the length of path searched multiplied by four meters (for searching two meters on either side of the central path line) and the area of the 10-meter radius circular site. We subtracted the overlap in the search path and site by subtracting four \* 10 meters.

*Web Position* -- To explore differences in web placement, we measured the web height from the ground to the web retreat for every web within the ten-meter radius. Differences in web height have been shown to vary by environmental factors associated with urbanization [@dahirel_2017].

We collected data from September 23 to October 14, 2020. During these dates, all *A. pennsylvanica* had reached maturity [personal observation, @guarisco_2014].

### Environmental predictors of spider spatial patterns

To assess potential environmental influences on variation in abundance, aggregation, and position, we looked for correlations between spider spatial patterns and the nine environmental conditions described earlier. In addition to exploring relationships across the data set overall, we subset the data by habitat to explore within habitats. We reduced the number of environmental predictors by removing variables that had a pairwise correlation above 0.6 using the *caret* package in R [@caret]. As a result of this analysis, artificial light at night radiance was dropped from all subsequent analyses. Only plant species richness, road disturbance, and highway disturbance remained for analyses of overall data. The urban center subset includes plant species richness, patch area, spectral radiance, road and highway disturbance, and road length. We kept tree cover, plant species richness, patch area, and road and highway disturbance for the urban forest subset.
